The Effects of the NBA COVID Bubble on the NBA Playoffs

### 2020-11-12 Manuscript/NBABubbleAnalysis.Rmd
Update NBABubbleAnalysis.Rmd

# Data {#sec:data}

\emph{need to give some descriptive statistics on how many games in
each sample}

Data were collected from the official NBA website. The main variables of interest
are whether or not the home team won, scoring totals for home and away teams, and
shooting percentages for home and away teams on two pointers, three pointers, and
free throws. These variables were very popular and frequently used in the related
literature discussed earlier. The data was collected on a game by game basis, giving us one observation
for each variable per game played. There were 83 games played in the 2020 playoffs. Giving us 83 observations 
for each variable in 2020 for both home and away teams. Likewise, there was 243 games played over 
2017-2019. Giving us 243 observations of each variable for both home and away teams over 2017-2020. 
While many other measures could be used for measuring
the outcome of the game and team performance, scoring seemed to be the most
important. The winner of a game is determined by who scores more points. So,
naturally when studying the outcome of winning, changes in scoring and scoring
efficiency are important to understand.

@@

A test for differences in means was conducted to compare scoring
totals. The null hypothesis is
\[
H_0: \mu_1 = \mu_2,
\]
where $\mu_1$ and $\mu_2$ are the mean scoring total in 2020 and
2017-2019, respectively. Let
$\bar{X}_1$ represent the average scoring total for 2020 and $\bar{X}_2$ represent
average scoring total over 2017-19. Similarly, let $s_1^2$ represent the sample variance
of scoring for 2020 and $s_2^2$ represent sample variance of scoring over 2017-19.
Finally, let $n_1$ represent the number of games in 2020 and $n_2$ represent the number
of games from 2017-2019. The null distribution of the test statistic is standard normal.
The test will be conducted twice once comparing home scoring in 2020 to 2017-19 and once
comparing away scoring in 2020 to 2017-2019. These tests will also be simultaneously
conducted using the Wilcoxon Test. The advantage of running the Wilcoxon Test is as a
nonparametric test it does not rely on any normality assumption

@@

Starting from the top of the 
Table \@ref(tab:table) summarizes results of the nine hypotheses
including the testing statistics and their p-values. Also reported are
the point estimates of the two samples in each comparison and the
confidence intervals of the true difference between the two.

First, we see a statistically significant change
in home win percentage in 2020 from 2017-19, with p-value of 0.0365. The 95%
confidence interval of (-0.255, -0.008) confirms our belief that home-court
advantage was lost in the 2020 NBA playoffs. However, after accounting for
multiple tests using the Bonferroni correction, the p-value is no longer significant.
So, we may only cautiously say there is evidence that home-court advantage was
not a factor in 2020.

Home team performance did not seem to be negatively impacted by losing home-court
advantage like expected. Home scoring, two point, three point shooting and free
throw shooting all show no significant difference, on average, between 2020 vs.
2017-19. There appear to be no evidence suggesting home teams played at
a lower level in 2020 than they did in previous years when they had home-court advantage.

\emph{focus on the two tests that are significant after Bonferroni
correction.}

Away teams saw more of an impact than home teams. For starters, there is a
significant increase in mean points per game by about 5 points (95% CI: 2.083, 7.988)
Likewise, the away team two point shooting efficiency increased by an average of
about 0.03 (95% CI: 0.015, 0.050). However, unlike two point shooting, away
teams did not see a statistically significant difference in three point and free
throw shooting. Overall, away teams have evidence of change in performance in
the bubble. The away teams seemed to perform better than they would under normal
conditions as a visiting team.

>>>>

# Data {#sec:data}
Data were collected from the official NBA website. The main variables of interest
are whether or not the home team won, scoring totals for home and away teams, and
shooting percentages for home and away teams on two pointers, three pointers, and
free throws. These variables were very popular and frequently used in the related
literature discussed earlier. The data was collected on a game by game basis, giving us one observation
for each variable per game played. There were 83 games played in the 2020 playoffs. Giving us 83 observations 
for each variable in 2020 for both home and away teams. Likewise, there was 243 games played over 
2017-2019. Giving us 243 observations of each variable for both home and away teams over 2017-2020. 
While many other measures could be used for measuring
the outcome of the game and team performance, scoring seemed to be the most
important. The winner of a game is determined by who scores more points. So,
naturally when studying the outcome of winning, changes in scoring and scoring
efficiency are important to understand.

@@

A test for differences in means was conducted to compare scoring
totals. The null hypothesis is
\[
H_0: \mu_1 = \mu_2,
\]
where $\mu_1$ and $\mu_2$ are the mean scoring total in 2020 and
2017-2019, respectively. Let
$\bar{X}_1$ represent the average scoring total for 2020 and $\bar{X}_2$ represent
average scoring total over 2017-19. Similarly, let $s_1^2$ represent the sample variance
of scoring for 2020 and $s_2^2$ represent sample variance of scoring over 2017-19.
Finally, let $n_1$ represent the number of games in 2020 and $n_2$ represent the number
of games from 2017-2019. The null distribution of the test statistic is standard normal.
The test will be conducted twice once comparing home scoring in 2020 to 2017-19 and once
comparing away scoring in 2020 to 2017-2019. These tests will also be simultaneously
conducted using the Wilcoxon Test. The advantage of running the Wilcoxon test is as a
nonparametric test it does not rely on any normality assumption

@@

Starting from the top of the 
Table \@ref(tab:table) summarizes p-values of the nine hypotheses for both
z-tests and Wilcoxon tests. The p-values are all fairly similar for both
tests giving strong confidence in conclusions drawn. Also reported are 
the point estimates of the two samples in each comparison.

First, we see a statistically significant change
in home win percentage in 2020 from 2017-19, with p-value of 0.0365 for the
z-test and 0.0369 for the Wilcoxon test. The 95% confidence interval of (-0.255, -0.008) 
confirms our belief that home-court advantage was lost in the 2020 NBA playoffs. However, after accounting for
multiple tests using the Bonferroni correction, the p-values for both tests are no longer significant.
So, we may only cautiously say there is evidence that home-court advantage was
not a factor in 2020.

Home team performance did not seem to be negatively impacted by losing home-court
advantage like expected. Home scoring, two-point and three-point shooting
shooting all show no significant difference, on average, between 2020 vs.
2017-19 based on p-values from both tests. However, the Wilcoxon test and z-test have conflicting results
for free throws. The z-test p-value of 0.0692 indictaes no significant difference, while
the Wilcoxon test p-value of 0.0469 indicates a difference at the 5% signficance level. Since,
the p-value of Wilcoxon test is so close to significance level and neither p-value is signifcant
after a Bonferonni correction for multiple tests this difference is likely not very meaningful.
There appears to be no strong evidence suggesting home teams played at
a lower level in 2020 than they did in previous years when they had home-court advantage.

Away teams saw more of an impact than home teams. For starters, there is a
significant increase in mean points per game, indicated by p-value of 0.0008 for z-test and
0.0004 for Wilcoxon. It is important to note both p-values also remain significant after
a Bonferonni correction giving strong indication of significance. The average difference in points is estimated to 
be about 5 points (95% CI: 2.083, 7.988).Likewise, the away team two point
shooting efficiency increased significantly based on p-value of 0.0003
for both the z-test and Wilcoxon test. Again both p-values remain signficant after Bonferroni correction.
The average difference can be estimated to be about 0.03 (95% CI: 0.015, 0.050). However,
unlike two point shooting, away teams did not see a statistically significant difference in three point and free
throw shooting. Overall, away teams have evidence of change in performance in
the bubble. The away teams seemed to perform better than they would under normal
conditions as a visiting team.
